Software Development in Startup Companies: The Green-field Startup Model

### 2013-03-04 article/main.tex
Fixed potential problems with references starting on a new line

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
%
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
%
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
%
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
%
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{A}{n} impressive number of new software startups are launched
worldwide every day as a result of an increase of new markets, accessible
technologies, and venture capital \cite{8491286}. With the term
\textit{software startups} we refer to those temporary organizations focused on
the creation of high-tech and innovative products, with little or no operating
history, aiming to grow by aggressively scaling their business in highly
scalable markets.

New ventures such as \textit{Facebook}, \textit{Linkedin}, \textit{Spotify},
\textit{Pinterest}, \textit{Instagram}, \textit{Groupon} and \textit{Dropbox},
to name a few, are good examples of startups that evolved into successful
businesses. But, despite many successful stories, the great majority of startups
fail within two years from their creation, primarily due to self-destruction
rather than competition \cite{Crowne2002}. Operating in a chaotic, rapidly
evolving and uncertain domain, software startups face intense time-pressure from
the market and are exposed to tough competition \cite{Maccormack2001,
Eisenhardt1998}. In order to succeed in this environment, it is crucial to
choose the right features to build and be able to quickly adapt the product to
new requests constrained by very limited amount of resources \cite{Sutton2000}.

From a software engineering perspective development in startups is challenging
since they develop software in a context where processes can hardly follow a
prescriptive methodology \cite{Sutton2000, Coleman2005}. Despite startups share
some characteristics with similar domains (e.g. small and web companies), the
combination of different factors makes the specific development context quite
unique \cite{Blank2005, Sutton2000}. Therefore, research is needed to support
their engineering activities \cite{Coleman2005}, guiding practitioners in taking
decisions and avoiding choices that could easily lead the whole business to
failure \cite{Kajko-Mattsson2008}. However, despite the impressive size of the
startup ecosystem  \cite{ISI:000243253000007}, the state-of-the-art presents a
gap. Through a  \textit{Systematic Mapping Study} (SMS) of the literature,
conducted by  Paternoster et al. \cite{SMS}, only few publications were found
related to  engineering activities in startups. Moreover the majority of these
studies do  not possess the relevance and rigour required to form a consistent
body of  knowledge in the state-of-the-art and in the state-of-practice.

@@

With a cross-sectional study we focused the research context in a limited  time-
frame to emphasize the nature of uncertainty in the development activities  and
time pressure from the market, as discussed in \cite{Maccormack2001,
Eisenhardt1998}. We performed semi-structured in-depth interviews with CEOs and
CTOs of startups, covering a wide spectrum of thematics and iteratively
adjusting the direction of the research according to the emerging evidences.

@@

Practitioners can use the GSM to apply presented strategies to speed-up the
development by taking in consideration the amount of technical debt accumulated
during the first early-stage period. In this regard, we identified several
commonalities between the issues related to software development in startups and
the research focused at studying technical debt \cite{Nugroho2011,Izurieta2012}.
On the other hand, researchers can take advantage from the GSM to  understand
how the technical debt influences the future growth of startup companies.

The remainder of this paper is structured as follows:  we provide relevant
information about startups in section \ref{backg}. Section \ref{resmet}
introduces the research questions and discusses the design and execution of the
research process. Results are discussed in section \ref{res:gsm}, which presents
the GSM. Section \ref{res:val}  compares results of this research to  literature
and frameworks presented in this field. Section \ref{valt} discusses  the
validity threats considered during this study. Section  \ref{sect:theory:impl}
discusses the most relevant implications of the GSM.  Finally, this paper
concludes in section \ref{conc}.

@@

Modern entrepreneurship has been boosted by the advent of the consumer internet
markets in the middle of the 1990s and culminated with the notorious  \textit
{dot-com bubble burst} of 2000 \cite{Perkins:1999:IBI:555126,
storey1982entrepreneurship}. Several years later, with the massification of the
internet and mobile devices, we are now assisting to an impressive proliferation
of software ventures - metaphorically referred as the \textit{startup bubble}.
The easy access to vast potential markets and the low cost of services
distribution are appealing condition for modern entrepreneurs \cite{Marmer2011}.
Inspired by success stories, a large number of software businesses are created
every day. However, the great majority of these companies fail within two years
from their creation \cite{Crowne2002}.

Looking at the number of new business incubators which appeared in the last
decade one can estimate the importance of startups \cite{Grimaldi2005}. The wave
of disruption in new technologies has led companies to be more and more
competitive, forcing themselves to radical organizational and innovational
renewals,  which bring many companies to the attempt of behaving like startups
\cite{Christensen1997}.

The implementation of methodologies to structure and control the development
activities in startups is a major challenge for engineers \cite{Coleman2008}.
Generally, the management of software development is achieved through the
introduction of software processes, which define what steps the development
organizations should take at each stage of production and provide assistance in
making estimates, developing plans and measuring quality \cite{Chroust1996}. In
the last decades, several models have been introduced to control software
development activities. However, their application in startup companies doesn't
report significant benefits \cite{Coleman2008a, Coleman2008, Sutton2000}.

In this context, software engineering (SE) faces complex and multifaceted
obstacles in understanding how to manage development processes. Sutton defines
startups as creative and flexible in nature and reluctant to introduce process
or bureaucratic measures which may result ineffective \cite{Sutton2000}. Also
Bach refers to startups as \textit{``a bunch of energetic and committed people
without defined development processes''} \cite{Bach1998}. In fact, startups have
very limited resources and typically wish to use them to support product
development instead of establishing processes \cite{Coleman2008, Heitlager2007}.
Some attempts to taylor lightweight processes to startups reported basic failure
of their application: \textit{``Everyone is busy, and software engineering
practices are often one of the first places developers cut corners''}
\cite{Martin2007}. Rejecting the notion of repeatable and controlled processes,
startups prominently take advantage of unpredictable, reactive and low-precision
\cite{surviving-os-cockburn} engineering practices \cite{Sutton2000,
Tanabian2005, Chorev2006, Kakati2003}.

Moreover, most startups develop packaged applications rather than software for
a specific client \cite{genome2012}. Issues related to this domain are addressed
in literature by the area known as \textit{market-driven software development}
(sometimes called \textit{packaged software development} or \textit{COTS
software development} \cite{Alves2006}). Among other results, researchers
emphasize the importance of time-to-market as a key strategic objective
\cite{dagMDR,sawyer99} for companies operating in this domain.  Other peculiar
aspects which influence the software development are related to requirements,
which are reported to be often ``invented by the software company''
\cite{512553}, ``rarely documented'' \cite{Karlsson02challengesin}, and can be
validated only after the product is released to market
\cite{dahl2003,Keil:1995}. Under these circumstances, failure of product
launches are largely due to ``products not meeting customer needs''
\cite{Alves2006}.

Accordingly, product-oriented practices help startups having flexible teams,
with workflows that leave them the ability to quickly change the direction
according to the targeted market  \cite{Heitlager2007,Sutton2000}. In this
regard, many startups focus on team productivity, asserting more control to the
employees instead of providing them rigid guidelines \cite{Tanabian2005,
Chorev2006, Kakati2003}.

%------------------------------------------------------------------------------%
\subsection{General lack of research in startups} Despite some studies tried to
address the above mentioned issues, a SMS  \cite{SMS} revealed only a few SE
works in this specific area, as confirmed by  other studies \cite{Coleman2008,
Coleman2008a, Coleman2007, Sutton2000}.  Moreover, the studies, identified in
\cite{SMS}, appear to be highly fragmented  and spread across different areas
rather than constituting a consistent  \textit{body of knowledge}.

A research published in 2000 by Sutton \cite{Sutton2000} confirmed a general
lack of  studies in this area, claiming that \textit{``software startups
represent a segment that has been mostly neglected in process studies''} and it
has been further confirmed with the empirical studies of Coleman et al.
\cite{Coleman2008,Coleman2008a,Coleman2007} eight years later.

@@

The word \textit{startup} appeared in the SE literature for the first time in
1994 in an article written by Carmel \cite{Camel1994a} where he studied the
time-to-completion in a young package firm. %MUN: If this is a quote  
%from Carmel, put it in quotation marks, like you did in the previous section  
%with Suttons quote. Generally, check that you are consistent how you use  
%quotations throughout the paper.  
%CG: fixed, btw  it was just to emphasize his the focus of his study.  
%MUN1: Ok. As you may have noticed, I have removed a lot of cases were you use  
%italics. If you emphasize *too* much, it it rather distracting than helpful  
%for the reader. Furthermore, it conflicts with you convention you use to  
%highlight high level categories of GSM. CG: ok. 
He noticed how these companies were particularly innovative and successful,  
advocating for the need of more research investigating software development  
practices so as to replicate success and try to transfer it to other 
technology sectors.

First insights reveal how software startups are product-oriented in the first
period of their development phase \cite{Heitlager2007}. Despite good
achievements at the beginning, software development and organizational
management increase in complexity \cite{1456074,Banker1998} causing
deterioration of performance over time. Briefly, the necessity of establishing
initial repeatable and scalable processes cannot be postponed forever.

A study of Kajko-Mattsson \cite{Kajko-Mattsson2008}, which investigated a
Swedish software startup, reported a heavy lack of requirements gathering
processes, minimal project management, lack of control over the change requests,
absence of documentation to track the status and progress of the process, and
defective releases. Accordingly, Ambler et al. report how two startups
approaching to an upcoming IPO started to require processes to focus on scalable
solutions, in view of the growing company's size in terms of users and employees
\cite{Ambler2002}. In this regard, Crowne \cite{Crowne2002} specifies different
stages through which software startups evolve. Starting without any established
workflows, startups grow over time, creating and stabilizing processes to
eventually improving them only when sufficiently mature.

But when startups have no time for training and orienting activities, as
discussed in \cite{Sutton2000}, their main focus remains on team capabilities
instead of prescriptive processes, hiring people who can \textit{``hit the
ground running''} \cite{Yoffie1999}. Empowering the team and focus on
methodological attributes of the processes oriented in prototyping,  proof-of-
concepts, mock-ups and demos, to test basic functionalities, have been  the
primary priority in startups \cite{Camel1994a}. Only when they grow, formal
methodologies arise, followed by performing quality assurance and long-term
planning processes \cite{Yoffie1999}.

The maturity of a company affects the extent to which processes are adopted.
Tingling \cite{Tingling2007} reports how introducing Extreme Programming (XP)
principles \cite{Beck:2004:EPE:1076267} in the development process was
challenging because of the need of trained team-members for fully implementing
the methodology. Da Silva and Kon \cite{Silva2005} were able to start with all
the XP practices in place only after six months of coaching the team, trying to
enhance maturity from day-one. Nevertheless, even then, customization of
practices were inevitably implemented to adapt the processes to the undertaken
startups' context \cite{Deias}.

As partially discussed above, contributions to flexibility and reactiveness of
the development process has been conducted prominently by means of \textit{Lean}
\cite{Gautam2008} and \textit{Agile} \cite{Abrahamsson2002} methodologies (also
reported in \cite{Taipale2010, Kuvinka2011}), where the extreme uncertain
conditions lead startups to learn fast from trials and errors with a strong
customer relationship in order to avoid wasting time in building wrong
functionalities and prevent rapidly exhaustion of resources \cite{Midler2008,
Hilmola2003, Sutton2000}. Customer involvement in software development has also
been discussed by Yogendra \cite{Yogendra2002} as an important factor to
encourage an early alignment of business concerns to the technology strategies,
because both are salient considerations to be successful \cite{Tingling2007}.

When startups take a development approach that is mainly product-oriented
rather than process-oriented, it is essential to have a flexible team, with a
workflow that helps them quickly to change direction according to the target
market \cite{Sutton2000}. In this regard, many startups have been focused on
team productivity, providing more control to the employees instead of providing
them rigid guidelines \cite{Tanabian2005, Chorev2006, Kakati2003}.

@@

When startups are mature enough to support software process improvement (SPI),
the solutions considered according to the state-of-the-art are oriented towards
light-weight processes such as a design of development process based on XP,
proposed by Zettel \cite{Zettel2001}. The process consists of a set of
activities and artifacts (in addition to some important roles) defined in order
to identify responsibilities and tools to utilize. But, despite the promising
benefits reported by Zettel, we were not able to identify any future evaluation
in real-world settings.

Another attempt of SPI in startups has been conducted by Deakins et al.
introducing a \textit{Helical Model for managing e-commerce development
environment} \cite{Deakins2005}. Also in this case, the author prescribed broad
guidelines for a rapid high-quality development process, which underwent limited
testing only in academic settings.

The problem of \textit{one-size-fits-all}, related to the SPI  representations
for startups, is described by Fayad \cite{Fayad1997}. The author discusses the
problem in actuating the same \textit{best-practices} criteria for established
companies in 10-person software startups. Thoroughly remarked in
\cite{Sutton2000}, Sutton states that problems of SPI in software startups arise
because of: the dynamism of the development process, which precludes
repeatability; organizational maturity, that cannot be maintained from startups
in view of lack of corporate direction; severe lack of resources, both human and
technological for process definition, implementation, management, and training.
In conclusion, the primary benefits of SPI do not address startups, which
instead of promoting product quality, aim to minimize time-to-market.

Additionally, the role of SPI has been neglected because seen as an obstacle to
the development team's creativity and flexibility, and to  the need of a quick
delivering product process environment \cite{Coleman2008a}. 
%MUN: the second part of the sentence is just a statement without reference. If 
%it refers to a statement by Coleman, put his reference at the end OR find the  
%correct reference OR delete the statement. %CG fixed 
Product quality is left aside in favor of minimal and suitable functionalities  
to shorten the time-to-market. As reported in two studies of Mater and Mirel  
\cite{Mater2000, Mirel2000}, quality aspects, mostly taken in consideration in  
internet startups, are oriented to usability and scalability, even though the  
market and application type heavily influences the quality-demand  
\cite{Coleman2008, Kim2005}.

To maintain the development activities, oriented to limited but suitable
functionality, studies suggest to externalize the complexity of parts of the
project to third party solutions by means of outsourcing activities
\cite{Hanna2010}, software reuse \cite{Jansen2008} and open-source strategies ,
\cite{Wall2001, Bean2005}.  %MUN: Can you split up the references? I.e. write:
%"outsourcing activities \cite{Bla}, software reuse \cite{Bla}, and open-source
%strategies \cite{Bla}. %CG: fixed

%------------------------------------------------------------------------------%
\subsection{Technical debt} Notwithstanding, a new area of SE research,
trying to tackle the problem of the \textit{technical debt}, brings and
encompasses different implications in studying development in software startups
\cite{Tom2013}. The metaphoric neologism was originally introduced by
Cunningham in 1992 \cite{TechnicalDebtCunn} and has recently attracted the
attention of SE researchers\footnote{To attest the fact that \textit{technical
debt} is gaining  traction among researchers, we mention two important
contributions which characterize the \textit{``debt landscape''}:
\cite{Nugroho2011,Izurieta2012} and a dedicated workshop \cite{workshopDebt}
organized by the Software  Engineering Institute  and ICSE.}. The concept of
\textit{technical debt} is  well illustrated in
\cite{Brown:2010:MTD:1882362.1882373}: \textit{``The idea  is that developers
sometimes accept compromises in a system in one dimension  (e.g., modularity) to
meet an urgent demand in some other dimension (e.g., a  deadline), and that such
compromises incur a ``debt'' on which ``interest'' has  to be paid and which the
``principal'' should be repaid at some point for the long-term health of the
project''}.  The compromise between high-speed and high-quality engineering is
faced daily by software startups, not only in terms  of architecture design but
in multifaceted aspects (weak project management, testing, process control).

Since \textit{``all decisions related to product development are trade-off
situations''} \cite{Hilmola2003}, generally startups optimize workflows to the
dynamic context they are involved into. They typically adopt any development
style that might work to support their first needs in what is called the
\textit{``Just do it''} school of software startups \cite{Ries2011}.
Additionally, as remarked by Coleman, \textit{``many managers just decide to
apply what they know, as their experience tells them it is merely common
sense''} \cite{Coleman2008}.

@@

\begin{compactitem} 
\item Software development strategy: the overall approach adopted by the  
company to carry out the product development. 
\item Engineering activities: the activities needed to bring a product from  
idea to market. Traditional engineering activities are, among others,  
requirement engineering, design, architecture, implementation, testing. 
\item Engineering elements: any practice, tool and artifacts contributing and  
supporting the engineering activities.
\item Quality attributes: those overall factors that affect run-time
behavior, system design, and user experience. They represent areas of concern
that have  the potential for applications to impact across layers and tiers.
Some of these  attributes are related to the overall system design, while others
are specific  to run time, design time, or user centric issues
\cite{Microsoft2009}. 
\item Growth: an increase in the company size respect to the initial 
conditions in terms of either employees or users/customers, and
product complexity in terms  of handling an increasing number of feature
requests. 
\item Software product: any software product and software service.
\end{compactitem}

@@

The aim of this research is to \textit{understand how software development
strategies are engineered by practitioners in startup companies in terms of
level of: structure, planning and control of software projects, in the period of
time  that goes from idea conception to the first open beta release of the
software product} \footnote{The above mentioned goal is structured according to
the GQM paradigm \cite{Basili1992}.}.

Initially the boundaries of the research domain were set by means of  non-
systematic literature survey, which provided the initial keywords used  further
in the research process. Moreover it helped us in defining the research
questions (RQs):

@@

We investigated the software development approach undertaken by practitioners
of startups. Following a \textit{Grounded Theory} (GT) methodology
\cite{Glaser1978}, we executed 13 semi-structured interviews integrated with
follow-up questionnaires. From this, we elaborated and extracted the Green-field
startup model (GSM) explaining the underlying phenomenon of software development
in startups.

>>>>

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
%
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
%
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
%
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
%
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{A}{n} impressive number of new software startups are launched
worldwide every day as a result of an increase of new markets, accessible
technologies, and venture capital~\cite{8491286}. With the term
\textit{software startups} we refer to those temporary organizations focused on
the creation of high-tech and innovative products, with little or no operating
history, aiming to grow by aggressively scaling their business in highly
scalable markets.

New ventures such as \textit{Facebook}, \textit{Linkedin}, \textit{Spotify},
\textit{Pinterest}, \textit{Instagram}, \textit{Groupon} and \textit{Dropbox},
to name a few, are good examples of startups that evolved into successful
businesses. But, despite many successful stories, the great majority of startups
fail within two years from their creation, primarily due to self-destruction
rather than competition~\cite{Crowne2002}. Operating in a chaotic, rapidly
evolving and uncertain domain, software startups face intense time-pressure from
the market and are exposed to tough competition~\cite{Maccormack2001,
Eisenhardt1998}. In order to succeed in this environment, it is crucial to
choose the right features to build and be able to quickly adapt the product to
new requests constrained by very limited amount of resources~\cite{Sutton2000}.

From a software engineering perspective development in startups is challenging
since they develop software in a context where processes can hardly follow a
prescriptive methodology~\cite{Sutton2000, Coleman2005}. Despite startups share
some characteristics with similar domains (e.g. small and web companies), the
combination of different factors makes the specific development context quite
unique~\cite{Blank2005, Sutton2000}. Therefore, research is needed to support
their engineering activities~\cite{Coleman2005}, guiding practitioners in taking
decisions and avoiding choices that could easily lead the whole business to
failure~\cite{Kajko-Mattsson2008}. However, despite the impressive size of the
startup ecosystem~\cite{ISI:000243253000007}, the state-of-the-art presents a
gap. Through a \textit{Systematic Mapping Study} (SMS) of the literature,
conducted by Paternoster et al.~\cite{SMS}, only few publications were found
related to engineering activities in startups. Moreover the majority of these
studies do not possess the relevance and rigour required to form a consistent
body of knowledge in the state-of-the-art and in the state-of-practice.

@@

With a cross-sectional study we focused the research context in a limited 
time-frame to emphasize the nature of uncertainty in the development activities 
and time pressure from the market, as discussed in~\cite{Maccormack2001,
Eisenhardt1998}. We performed semi-structured in-depth interviews with CEOs and
CTOs of startups, covering a wide spectrum of thematics and iteratively
adjusting the direction of the research according to the emerging evidences.

@@

Practitioners can use the GSM to apply presented strategies to speed-up the
development by taking in consideration the amount of technical debt accumulated
during the first early-stage period. In this regard, we identified several
commonalities between the issues related to software development in startups and
the research focused at studying technical debt~\cite{Nugroho2011,Izurieta2012}.
On the other hand, researchers can take advantage from the GSM to understand
how the technical debt influences the future growth of startup companies.

The remainder of this paper is structured as follows: we provide relevant
information about startups in section~\ref{backg}. Section~\ref{resmet}
introduces the research questions and discusses the design and execution of the
research process. Results are discussed in section~\ref{res:gsm}, which presents
the GSM. Section~\ref{res:val} compares results of this research to literature
and frameworks presented in this field. Section~\ref{valt} discusses the
validity threats considered during this study. Section~\ref{sect:theory:impl}
discusses the most relevant implications of the GSM. Finally, this paper
concludes in section~\ref{conc}.

@@

Modern entrepreneurship has been boosted by the advent of the consumer internet
markets in the middle of the 1990s and culminated with the notorious \textit
{dot-com bubble burst} of 2000~\cite{Perkins:1999:IBI:555126,
storey1982entrepreneurship}. Several years later, with the massification of the
internet and mobile devices, we are now assisting to an impressive proliferation
of software ventures - metaphorically referred as the \textit{startup bubble}.
The easy access to vast potential markets and the low cost of services
distribution are appealing condition for modern entrepreneurs~\cite{Marmer2011}.
Inspired by success stories, a large number of software businesses are created
every day. However, the great majority of these companies fail within two years
from their creation~\cite{Crowne2002}.

Looking at the number of new business incubators which appeared in the last
decade one can estimate the importance of startups~\cite{Grimaldi2005}. The wave
of disruption in new technologies has led companies to be more and more
competitive, forcing themselves to radical organizational and innovational
renewals, which bring many companies to the attempt of behaving like 
startups~\cite{Christensen1997}.

The implementation of methodologies to structure and control the development
activities in startups is a major challenge for engineers~\cite{Coleman2008}.
Generally, the management of software development is achieved through the
introduction of software processes, which define what steps the development
organizations should take at each stage of production and provide assistance in
making estimates, developing plans and measuring quality~\cite{Chroust1996}. In
the last decades, several models have been introduced to control software
development activities. However, their application in startup companies doesn't
report significant benefits~\cite{Coleman2008a, Coleman2008, Sutton2000}.

In this context, software engineering (SE) faces complex and multifaceted
obstacles in understanding how to manage development processes. Sutton defines
startups as creative and flexible in nature and reluctant to introduce process
or bureaucratic measures which may result ineffective~\cite{Sutton2000}. Also
Bach refers to startups as \textit{``a bunch of energetic and committed people
without defined development processes''}~\cite{Bach1998}. In fact, startups have
very limited resources and typically wish to use them to support product
development instead of establishing processes~\cite{Coleman2008, Heitlager2007}.
Some attempts to taylor lightweight processes to startups reported basic failure
of their application: \textit{``Everyone is busy, and software engineering
practices are often one of the first places developers cut 
corners''}~\cite{Martin2007}. Rejecting the notion of repeatable and controlled 
processes, startups prominently take advantage of unpredictable, reactive and 
low-precision~\cite{surviving-os-cockburn} engineering 
practices~\cite{Sutton2000,Tanabian2005, Chorev2006, Kakati2003}.

Moreover, most startups develop packaged applications rather than software for
a specific client~\cite{genome2012}. Issues related to this domain are addressed
in literature by the area known as \textit{market-driven software development}
(sometimes called \textit{packaged software development} or \textit{COTS
software development}~\cite{Alves2006}). Among other results, researchers
emphasize the importance of time-to-market as a key strategic 
objective~\cite{dagMDR,sawyer99} for companies operating in this domain. Other 
peculiar aspects which influence the software development are related to 
requirements, which are reported to be often ``invented by the software 
company''~\cite{512553}, ``rarely documented''~\cite{Karlsson02challengesin}, 
and can be validated only after the product is released to 
market~\cite{dahl2003,Keil:1995}. Under these circumstances, failure of product
launches are largely due to ``products not meeting customer 
needs''~\cite{Alves2006}.

Accordingly, product-oriented practices help startups having flexible teams,
with workflows that leave them the ability to quickly change the direction
according to the targeted market~\cite{Heitlager2007,Sutton2000}. In this
regard, many startups focus on team productivity, asserting more control to the
employees instead of providing them rigid guidelines~\cite{Tanabian2005,
Chorev2006, Kakati2003}.

%------------------------------------------------------------------------------%
\subsection{General lack of research in startups} Despite some studies tried to
address the above mentioned issues, a SMS~\cite{SMS} revealed only a few SE
works in this specific area, as confirmed by other studies~\cite{Coleman2008,
Coleman2008a, Coleman2007, Sutton2000}. Moreover, the studies, identified 
in~\cite{SMS}, appear to be highly fragmented and spread across different areas
rather than constituting a consistent \textit{body of knowledge}.

A research published in 2000 by Sutton~\cite{Sutton2000} confirmed a general
lack of studies in this area, claiming that \textit{``software startups
represent a segment that has been mostly neglected in process studies''} and it
has been further confirmed with the empirical studies of Coleman et 
al.~\cite{Coleman2008,Coleman2008a,Coleman2007} eight years later.

@@

The word \textit{startup} appeared in the SE literature for the first time in
1994 in an article written by Carmel~\cite{Camel1994a} where he studied the
time-to-completion in a young package firm. %MUN: If this is a quote  
%from Carmel, put it in quotation marks, like you did in the previous section  
%with Suttons quote. Generally, check that you are consistent how you use  
%quotations throughout the paper.  
%CG: fixed, btw  it was just to emphasize his the focus of his study.  
%MUN1: Ok. As you may have noticed, I have removed a lot of cases were you use  
%italics. If you emphasize *too* much, it it rather distracting than helpful  
%for the reader. Furthermore, it conflicts with you convention you use to  
%highlight high level categories of GSM. CG: ok. 
He noticed how these companies were particularly innovative and successful, 
advocating for the need of more research investigating software development 
practices so as to replicate success and try to transfer it to other 
technology sectors.

First insights reveal how software startups are product-oriented in the first
period of their development phase~\cite{Heitlager2007}. Despite good
achievements at the beginning, software development and organizational
management increase in complexity~\cite{1456074,Banker1998} causing
deterioration of performance over time. Briefly, the necessity of establishing
initial repeatable and scalable processes cannot be postponed forever.

A study of Kajko-Mattsson~\cite{Kajko-Mattsson2008}, which investigated a
Swedish software startup, reported a heavy lack of requirements gathering
processes, minimal project management, lack of control over the change requests,
absence of documentation to track the status and progress of the process, and
defective releases. Accordingly, Ambler et al. report how two startups
approaching to an upcoming IPO started to require processes to focus on scalable
solutions, in view of the growing company's size in terms of users and 
employees~\cite{Ambler2002}. In this regard, Crowne~\cite{Crowne2002} specifies 
different stages through which software startups evolve. Starting without any 
established workflows, startups grow over time, creating and stabilizing 
processes to eventually improving them only when sufficiently mature.

But when startups have no time for training and orienting activities, as
discussed in~\cite{Sutton2000}, their main focus remains on team capabilities
instead of prescriptive processes, hiring people who can \textit{``hit the
ground running''}~\cite{Yoffie1999}. Empowering the team and focus on
methodological attributes of the processes oriented in prototyping, 
proof-of-concepts, mock-ups and demos, to test basic functionalities, have been  
the primary priority in startups~\cite{Camel1994a}. Only when they grow, formal
methodologies arise, followed by performing quality assurance and long-term
planning processes~\cite{Yoffie1999}.

The maturity of a company affects the extent to which processes are adopted.
Tingling~\cite{Tingling2007} reports how introducing Extreme Programming (XP)
principles~\cite{Beck:2004:EPE:1076267} in the development process was
challenging because of the need of trained team-members for fully implementing
the methodology. Da Silva and Kon~\cite{Silva2005} were able to start with all
the XP practices in place only after six months of coaching the team, trying to
enhance maturity from day-one. Nevertheless, even then, customization of
practices were inevitably implemented to adapt the processes to the undertaken
startups' context~\cite{Deias}.

As partially discussed above, contributions to flexibility and reactiveness of
the development process has been conducted prominently by means of 
\textit{Lean}~\cite{Gautam2008} and \textit{Agile}~\cite{Abrahamsson2002} 
methodologies (also reported in~\cite{Taipale2010, Kuvinka2011}), where the 
extreme uncertain conditions lead startups to learn fast from trials and errors 
with a strong customer relationship in order to avoid wasting time in building 
wrong functionalities and prevent rapidly exhaustion of 
resources~\cite{Midler2008,Hilmola2003, Sutton2000}. Customer involvement in 
software development has also been discussed by Yogendra~\cite{Yogendra2002} as 
an important factor to encourage an early alignment of business concerns to the 
technology strategies, because both are salient considerations to be 
successful~\cite{Tingling2007}.

When startups take a development approach that is mainly product-oriented
rather than process-oriented, it is essential to have a flexible team, with a
workflow that helps them quickly to change direction according to the target
market~\cite{Sutton2000}. In this regard, many startups have been focused on
team productivity, providing more control to the employees instead of providing
them rigid guidelines~\cite{Tanabian2005, Chorev2006, Kakati2003}.

@@

When startups are mature enough to support software process improvement (SPI),
the solutions considered according to the state-of-the-art are oriented towards
light-weight processes such as a design of development process based on XP,
proposed by Zettel~\cite{Zettel2001}. The process consists of a set of
activities and artifacts (in addition to some important roles) defined in order
to identify responsibilities and tools to utilize. But, despite the promising
benefits reported by Zettel, we were not able to identify any future evaluation
in real-world settings.

Another attempt of SPI in startups has been conducted by Deakins et al.
introducing a \textit{Helical Model for managing e-commerce development
environment}~\cite{Deakins2005}. Also in this case, the author prescribed broad
guidelines for a rapid high-quality development process, which underwent limited
testing only in academic settings.

The problem of \textit{one-size-fits-all}, related to the SPI representations
for startups, is described by Fayad~\cite{Fayad1997}. The author discusses the
problem in actuating the same \textit{best-practices} criteria for established
companies in 10-person software startups. Thoroughly remarked 
in~\cite{Sutton2000}, Sutton states that problems of SPI in software startups 
arise because of: the dynamism of the development process, which precludes
repeatability; organizational maturity, that cannot be maintained from startups
in view of lack of corporate direction; severe lack of resources, both human and
technological for process definition, implementation, management, and training.
In conclusion, the primary benefits of SPI do not address startups, which
instead of promoting product quality, aim to minimize time-to-market.

Additionally, the role of SPI has been neglected because seen as an obstacle to
the development team's creativity and flexibility, and to the need of a quick
delivering product process environment~\cite{Coleman2008a}. 
%MUN: the second part of the sentence is just a statement without reference. If 
%it refers to a statement by Coleman, put his reference at the end OR find the  
%correct reference OR delete the statement. %CG fixed 
Product quality is left aside in favor of minimal and suitable functionalities 
to shorten the time-to-market. As reported in two studies of Mater and 
Mirel~\cite{Mater2000, Mirel2000}, quality aspects, mostly taken in 
consideration in internet startups, are oriented to usability and scalability, 
even though the market and application type heavily influences the 
quality-demand~\cite{Coleman2008, Kim2005}.

To maintain the development activities, oriented to limited but suitable
functionality, studies suggest to externalize the complexity of parts of the
project to third party solutions by means of outsourcing 
activities~\cite{Hanna2010}, software reuse~\cite{Jansen2008} and open-source 
strategies~\cite{Wall2001, Bean2005}. %MUN: Can you split up the references? 
%I.e. write:"outsourcing activities \cite{Bla}, software reuse \cite{Bla}, and 
%open-source strategies \cite{Bla}. %CG: fixed

%------------------------------------------------------------------------------%
\subsection{Technical debt} Notwithstanding, a new area of SE research,
trying to tackle the problem of the \textit{technical debt}, brings and
encompasses different implications in studying development in software 
startups~\cite{Tom2013}. The metaphoric neologism was originally introduced by
Cunningham in 1992~\cite{TechnicalDebtCunn} and has recently attracted the
attention of SE researchers\footnote{To attest the fact that \textit{technical
debt} is gaining traction among researchers, we mention two important
contributions which characterize the \textit{``debt landscape''}:
\cite{Nugroho2011,Izurieta2012} and a dedicated workshop~\cite{workshopDebt}
organized by the Software Engineering Institute and ICSE.}. The concept of
\textit{technical debt} is well illustrated 
in~\cite{Brown:2010:MTD:1882362.1882373}: \textit{``The idea is that developers
sometimes accept compromises in a system in one dimension (e.g., modularity) to
meet an urgent demand in some other dimension (e.g., a deadline), and that such
compromises incur a ``debt'' on which ``interest'' has to be paid and which the
``principal'' should be repaid at some point for the long-term health of the
project''}. The compromise between high-speed and high-quality engineering is
faced daily by software startups, not only in terms of architecture design but
in multifaceted aspects (weak project management, testing, process control).

Since \textit{``all decisions related to product development are trade-off
situations''}~\cite{Hilmola2003}, generally startups optimize workflows to the
dynamic context they are involved into. They typically adopt any development
style that might work to support their first needs in what is called the
\textit{``Just do it''} school of software startups~\cite{Ries2011}.
Additionally, as remarked by Coleman, \textit{``many managers just decide to
apply what they know, as their experience tells them it is merely common
sense''}~\cite{Coleman2008}.

@@

\begin{compactitem} 
\item Software development strategy: the overall approach adopted by the 
company to carry out the product development. 
\item Engineering activities: the activities needed to bring a product from 
idea to market. Traditional engineering activities are, among others, 
requirement engineering, design, architecture, implementation, testing. 
\item Engineering elements: any practice, tool and artifacts contributing and 
supporting the engineering activities.
\item Quality attributes: those overall factors that affect run-time
behavior, system design, and user experience. They represent areas of concern
that have the potential for applications to impact across layers and tiers.
Some of these attributes are related to the overall system design, while others
are specific to run time, design time, or user centric 
issues~\cite{Microsoft2009}. 
\item Growth: an increase in the company size respect to the initial 
conditions in terms of either employees or users/customers, and
product complexity in terms of handling an increasing number of feature
requests. 
\item Software product: any software product and software service.
\end{compactitem}

@@

The aim of this research is to \textit{understand how software development
strategies are engineered by practitioners in startup companies in terms of
level of: structure, planning and control of software projects, in the period of
time that goes from idea conception to the first open beta release of the
software product}\footnote{The above mentioned goal is structured according to
the GQM paradigm~\cite{Basili1992}.}.

Initially the boundaries of the research domain were set by means of 
non-systematic literature survey, which provided the initial keywords used  
further in the research process. Moreover it helped us in defining the research
questions (RQs):

@@

We investigated the software development approach undertaken by practitioners
of startups. Following a \textit{Grounded Theory} (GT) 
methodology~\cite{Glaser1978}, we executed 13 semi-structured interviews 
integrated with follow-up questionnaires. From this, we elaborated and 
extracted the Green-field startup model (GSM) explaining the underlying 
phenomenon of software development in startups.

\textit{Grounded Theory} methodology is described by Robson as \textit{``the
best approach to answer the question - What is going on here?''} and defined by
its creators~\cite{ColinRobson2009} as \textit{``a set of well-developed
categories ( e.g. themes, concepts ) that are systematically interrelated
through statements of relations to form a theoretical framework that explains
some relevant social, psychological, educational, nursing or other
phenomenon''}.
